---
title: "Polynomial Class"
author: "Ryan Hastings, Mishek Thapa"
date: "May 27th, 2020"
output: html_document
---

Assignment for STA 323: Statistical Programming

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```

In this assignment, we create a class of polynomials using the `S3` object-oriented programming environment in `R`. A polynomial is 
constructed from two numeric vectors, one of coefficients and another of their 
respective powers.


### Class Description
We created a class of polynomials. A polynomial is constructed from a numeric 
vector of coefficients and uses its two attributes: the class, polynomial 
and powers, a numeric vector of each coefficient's respective power. The helper 
function here converts inputs of strings into numerical vectors for the user and
helps uncover some invalid inputs by creating NAs.

These two components of the constructor were supplemented with the validity 
checks in the validator function. The inputs that were filtered for: vectors 
with NA values, infinite values, and inputs that are not numbers. This step 
is significant because it gets rid of any non-real number inputs that could 
have been entered (characters, symbols, imaginary numbers). Another validity 
check was the check to make sure each vector (coefficient and power) were 
the same length.

### Functions
Our constructor function:

```{r constructor}
new_polynomial <- function(x = double(), y = double()) {
  
  # assign attributes to coefficient vector x
  structure(x, 
            powers = y,
            class = "polynomial")
}
```


Our validator function: 

```{r validator}
validate_polynomial <- function(x) {
  
  values <- unclass(x) #get the underlying base classes
  pow <-  attr(values,"powers")
  
  if (any(is.na(values)) | any(is.nan(values)) | any(is.infinite(values))) {
    stop("Coefficients are either infinite, not a number, or not available",call. = FALSE)
  }
  
  if (any(is.na(pow)) | any(is.nan(pow)) | any(is.infinite(pow))) {
    stop("Powers are either infinite, not a number, or not available", call. = FALSE)
  }
  
  if(length(x) != length(attr(x,"powers"))) {
    stop("The quantity of coefficients must be the same as the quantity of powers",call. = FALSE)
  }
  
  return(x)
}
```


Our helper function:

```{r helper}
polynomial <- function(x, y = 0:(length(x)-1)){
  
  # coerce inputs to type numeric
  x <- suppressWarnings(as.numeric(x))
  y <- suppressWarnings(as.numeric(y))
  
  # input into validator and constructor
  validate_polynomial(new_polynomial(x, y))
  
}
```

### Methods

The combination of these two components of the class allow for building upon
a few generic functions, specifically the print and the plot function. These 
two functions cycled through the two vectors to produce their output. First, the 
print function cycles through the elements of each vector in order and writes
one element of the expression and uses the paste function to combine the compo-
nents of the equations. The plot function creates a domain of x values to compute 
into the polynomial (can be defined by user), then cycles through each element
of the polynomial (like the print function) and adds them to compute each value
in the function, then plots all the points. 

The print function itself does not have any validation stop points because the 
polynomial function already has them, but the plot function includes a few additional.
Variables that specify the domain of the plot (lower, upper, step) that are infinite, not a 
number, or non-numeric are excluded inside this function. 


Our `print` method:

```{r print-polynomial}
print.polynomial <- function(x, ...) {
  
  # create a string displaying coefficients and their powers
  str <- paste(as.character(x), 
               " * x ^ ", 
               as.character(attr(x, "powers")),
               sep = "", 
               collapse = " + ")
  
  str <- paste("f(x) = ", str)
  print(str, ...)
}
```


Our `plot` method:

```{r plot-polynomial}  
plot.polynomial <- function(poly, 
                            lower = -5, # lower bound of region displayed
                            upper =  5, # upper bound of region displayed
                            step  = .1, # distance between each point evaluated
                            main  = "Polynomial Plot",
                            ylab  = "f(x)",
                            sub = print(poly),
                            type = "l",
                            ...) {
  
  #validation; stopifnot produces interpretable messages here
  stopifnot(attr(poly, "class") == "polynomial",
            is.numeric(lower),
            is.numeric(upper),
            is.numeric(step),
            !is.infinite(lower),
            !is.infinite(upper),
            !is.nan(lower),
            !is.nan(upper))
  
  # create function to evaluate a polynomial at a point
  eval_poly <- function(poly, num) {
  
    value <- 0
    
    for (i in 1:length(poly)) {
      value <- value + poly[i] * (num ^ (attr(poly, "powers")[i]))
    }
    
    return(value)
    }
  # generate function inputs and outputs to be plotted
  x <- seq(lower, upper, step)
  y <- eval_poly(poly, x)
  
  # plot using specified settings
  plot(x, y, main = main, ylab = ylab, sub = sub, type = type, ...)
}
```

### Usability and Robustness
By building on these generic functions with this class, the user can also use optional
features for each of these functions like different types of plotting styles for example.
This makes the function very robust and easy to use. A downside of this 
function is that the function takes in coefficients and powers separately. This
is not a very typical way of representing a polynomial, but it permits an efficient 
way to process the inputs. After understanding the input format though, the methods 
this class allows for are useful and very easy to use.  

Examples:

```{r valid_inputs, error = TRUE}
m <- polynomial(1, 2)
attributes(m)
print(m)
plot(m)
j <- polynomial(1:5, 5:9)
print(j)
plot(j)
n <- polynomial(pi)
print(n)
plot(n, lower = 0, upper = 8, main = "Pi", type = "p")
k <- polynomial(c(1, 3, 4, 2), c(-2:0, 1.4)) 
print(k)
plot(k, lower = 0.1, step = 0.1)
a <- polynomial(0.5, -1)
print(a)
plot(a, step = 0.1)
```







